# {{< fa broom size=1x >}} &nbsp;&nbsp;Data cleaning (15 min)

## Column names

R package `janitor` is of great help !

<br><br><br>

```{r}
#| echo: true

# Create a data.frame with dirty names

test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")


test_df |>
  janitor::clean_names() |>
  colnames()
```


## Sorting and arranging

<br>

R package `dplyr` is very useful to :

<br>

- **select** columns (based on their names only, no $var [,"var"] etc) : `dplyr::select(column1, column2)`


- **filter** specific lines from those selected columns : `dplyr::filter(level1 == a, level2 > x)`


- apply a function to the table subset : `group_by(factor_column)` and then `mutate(new_column = old_column*2)`


- put it back together in a dataframe/data table etc.


## dplyr demo

<br><br>

```{r}
#| echo: true

# load the iris dataset implemented in base R:
data("iris")
head(iris)
```

<br><br>

```{r}
#| echo: true

iris_nice <- janitor::clean_names(iris)
colnames(iris_nice)
```


## dplyr demo

<br><br>


```{r}
#| echo: true
iris_nice |>
  dplyr::select(sepal_length, sepal_width, species) |>
  dplyr::filter(sepal_length <= 5) |>
  dplyr::group_by(species) |>
  dplyr::mutate(mean_length = mean(sepal_length),
                mean_width = mean(sepal_width))

```

## Merging dataframes

The family of `dplyr::join_` functions can deal with all types of dataframe merges. It's the R equivalent to SQL querying. 

<br><br>

Load datasets from dplyr:

:::: {.columns}
::: {.column width=45%}

```{r}
#| echo: true

instruments <- dplyr::band_instruments |> 
  as.data.frame()
instruments
```
:::

::: {.column width=10%}
:::

::: {.column width=45%}

```{r}
#| echo: true

members <- dplyr::band_members |>
  as.data.frame()
members
```
:::
::::


## Merging dataframes

:::: {.columns}
::: {.column width=45%}

**`inner_join()`** only keeps observations from x that have a matching key in y.

```{r}
#| echo: true
#| warning: false


dplyr::inner_join(members, instruments) 
```

<br>


**`left_join()`** returns all x rows.
This is probably the most used one.

```{r}
#| echo: true
#| warning: false

dplyr::left_join(members, instruments) 
```

:::

::: {.column width=10%}
:::

::: {.column width=45%}

**`right_join()`** returns matched of x rows, followed by unmatched y rows.

```{r}
#| echo: true
#| warning: false

dplyr::right_join(members, instruments) 
```

<br>


**`full_join()`** returns all x rows, followed by unmatched y rows. Most conservative.

```{r}
#| echo: true
#| warning: false

dplyr::full_join(members, instruments) 
```

:::

::::



## Pivoting with tidyr

`pivot_longer()` : compiles information from multiple columns into a unique column. This is commonly needed to reformat datasets that were optimized for ease of data entry or comparison rather than ease of analysis. 

> This format is the one privileged by the ggplot functions.

```{r}
#| echo: true

relig_income <- tidyr::relig_income 
head(relig_income)
```


## Pivoting with tidyr

<br> 

:::: {.columns}
::: {.column width=45%}


```{r}
#| echo: true

newtab <- tidyr::pivot_longer(
  relig_income, 
  cols = !religion,
  names_to = "income",
  values_to = "count"
)
```

<br>

```{r}
#| echo: true
#| eval: false
head(newtab)
```

:::

::: {.column width=10%}
:::

::: {.column width=45%}


<br> <br> 


```{r}
#| echo: false
#| eval: true

head(newtab |> as.data.frame(), 12)
```

:::

::::






## Pivoting with tidyr

`pivot_wider()` is the opposite of `pivot_longer()`. <br> <br> 


```{r}
#| echo: true
back <- tidyr::pivot_wider(
  newtab,
  names_from = income,
  values_from = count
)
```

<br> <br> 

```{r}
#| echo: false
#| eval: true
head(back)

```


## Working on strings

```{r}
#| eval: false
#| echo: true

string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```

<br>

```{r}
#| eval: false
#| echo: true

> "This is a string without a closing quote
+ 
+ 
+ HELP I'M STUCK
```

<br> 

**`paste()`**: stick 2 strings together: 

```{r}
#| eval: true
#| echo: true
a <- "apple"
b <- "banana"
paste(a, b)
```
or collapse elements from a vector into a string: 

```{r}
#| eval: true
#| echo: true
(team <- c("Nico", "Camille", "Olivier", "Yvan", "Sophie", "Anne-Sophie"))
(paste(team, collapse = " - "))
```


## Working on strings

**Regular expressions**, or "regexps" : a concise language for describing patterns in strings. More [here](https://r4ds.had.co.nz/strings.html).

The `stringr` package uses a standardized set of functions applied to strings to:

- **detect matches:** `str_detect(string, pattern)` returns TRUE/FALSE, `str_which()` returns position in string (like `grep()` in base R), `str_count()` returns # of matches

- **replace matches:** `str_replace_all(string, pattern, replacement)` <--> same as `gsub()` in base R, useful to replace typos

- **subset matches:** `str_subset(string, pattern)` <--> same as `grep(patter, string, value=TRUE)` in base R

```{r}
#| eval: true
#| echo: true
stringr::str_which(team, "Ca")
stringr::str_subset(team, "Ca")
```
- **manipulate:** `str_flatten()` <--> same as `paste()`, `str_split(string, pattern)`: cuts a string where pattern matches.


- **trim white spaces:** `str_trim()` : very usefull!! Good to run just in case.... 

- **change character case:** `str_to_upper()`: ALL CAPS.



## Working on dates... and times

We recommend the `lubridate`  package, and its precious [cheatsheet](https://evoldyn.gitlab.io/evomics-2018/ref-sheets/R_lubridate.pdf).

### Parsing

Identify the order of the year (y), month (m), day (d), hour (h), minute (m) and second (s) elements in your data


```{r}
#| echo: true
#| warning: false

library(lubridate)

# all of these are equivalent: 
c(ymd("20110604"), mdy("06-04-2011"), dmy("04/06/2011"), mdy("June 4th, 2011"))

# date and time: you can specify the timezone:
 ymd_hms("2017-11-28T14:02:00", tz="Pacific/Auckland")
```
<br>

A date-time is a point on the timeline, stored as the number of seconds since 1970-01-01 00:00:00 UTC

```{r}
#| echo: true
as_datetime(1511870400)
```

<br> 

```{r}
#| echo: true

as_date(198660)

```




## New slide

- Packages `janitor`, `dplyr`, `lubridate`, `stringr`, `tidyr`

<!-- new list -->

- One slide on `janitor` and `dplyr` # done
- One or two slides on `*_join()` # done
- Long vs. tidy format w/ `tidyr` # done
- Working on strings w/ `stringr`, `paste()`, `strsplit()`, `gsub()`
- Working on dates w/ `lubridate`
