# {{< fa broom size=1x >}} &nbsp;&nbsp;Data cleaning (15 min)

## Column names

R package `janitor` is of great help !

```{r}
#| echo: true

# Create a data.frame with dirty names
test_df <- as.data.frame(matrix(ncol = 6))
names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")
```

```{r}
#| echo: true

test_df |>
  janitor::clean_names() |>
  colnames()
```

## Sorting and arranging

R package `dplyr` is very useful to :

- select columns (based on their names only, no $var [,"var"] etc) : `dplyr::select(column1, column2)`

- filter specific lines from those selected columns : `dplyr::filter(level1 == a, level2 > x)`

- apply a function to the table subset : `group_by(factor_column)` and then `mutate(new_column = old_column*2)`

- put it back together in a dataframe/data table etc.


## Demo

```{r}
#| echo: true

# load the iris dataset implemented in base R:
data("iris")
head(iris)
```


```{r}
#| echo: true

iris_nice <- janitor::clean_names(iris)
colnames(iris_nice)
```


## Demo

```{r}
#| echo: true
iris_nice |>
  dplyr::select(sepal_length, sepal_width, species) |>
  dplyr::filter(sepal_length <= 5) |>
  dplyr::group_by(species) |>
  dplyr::mutate(mean_length = mean(sepal_length),
                mean_width = mean(sepal_width))

```

## Merging dataframes

The family of `dplyr::join_` functions can deal with all types of dataframe merges. It's the R equivalent to SQL querying. 


```{r}
#| echo: true

# load datasets from dplyr
instruments <- dplyr::band_instruments |> as.data.frame()
instruments

members <- dplyr::band_members |> as.data.frame()
members
```


## Merging dataframes

**`inner_join()`** only keeps observations from x that have a matching key in y.

```{r}
#| warning: false

dplyr::inner_join(members, instruments) 
```

**`left_join()`** returns all x rows.
This is probably the most used one.

```{r}
#| warning: false

dplyr::left_join(members, instruments) 
```


## Merging dataframes

**`right_join()`** returns matched of x rows, followed by unmatched y rows.

```{r}
#| warning: false

dplyr::right_join(members, instruments) 
```

**`full_join()`** returns all x rows, followed by unmatched y rows.
Most conservative.
```{r}
#| warning: false

dplyr::full_join(members, instruments) 
```



- Packages `janitor`, `dplyr`, `lubridate`, `stringr`, `tidyr`

<!-- new list -->

- One slide on `janitor` and `dplyr` # done
- One or two slides on `*_join()` # done
- Long vs. tidy format w/ `tidyr`
- Working on strings w/ `stringr`, `paste()`, `strsplit()`, `gsub()`
- Working on dates w/ `lubridate`
