# {{< fa user-clock size=1x >}} &nbsp;&nbsp;Exercise (20 min)

## Cleaning data

**Part 1** - Clean the [PanTHERIA database](https://esajournals.onlinelibrary.wiley.com/doi/10.1890/08-1494.1) from the [previous exercise](https://biodiversitydata.github.io/data-access/#/accessing-data)

- use the function `readr::read_delim()` to import data
- select columns `MSW05_Binomial`, `5-1_AdultBodyMass_g`, `8-1_AdultForearmLen_mm` & `3-1_AgeatFirstBirth_d`
- rename columns in `pan_binomial_name`, `pan_adult_body_mass_g`, `pan_adult_forearm_len_mm` & `pan_age_at_first_birth_d`
- replace `-999` by `NA`

<br/>


**Part 2** - Create a reference table from GBIF occurrences downloaded [previously](https://biodiversitydata.github.io/data-access/#/accessing-data)

- remove all columns except `scientificName`
- remove duplicated rows
- clean binomial names (remove authors)

<br/>


**Part 3** - Merge this table with PanTHERIA (cleaned) to add trait values to these two species



## Correction


:::: {.columns}

::: {.column width=50%}
**Part 1**

```{r}
#| echo: true
#| eval: false

library(dplyr)

## Open PanTHERIA data ----

pantheria <- readr::read_delim("PanTHERIA_1-0_WR05_Aug2008.txt")


## Clean PanTHERIA data ----

pantheria <- pantheria |> 
  
  select(MSW05_Binomial, 
         `5-1_AdultBodyMass_g`, 
         `8-1_AdultForearmLen_mm`, 
         `3-1_AgeatFirstBirth_d`) |> 
  
  rename(binomial_name            = MSW05_Binomial,
         pan_adult_body_mass_g    = `5-1_AdultBodyMass_g`, 
         pan_adult_forearm_len_mm = `8-1_AdultForearmLen_mm`, 
         pan_age_at_first_birth_d = `3-1_AgeatFirstBirth_d`) |> 
  
  mutate(across(starts_with("pan_"), ~ ifelse(.x == -999, NA, .x)))
```
:::

::: {.column width=50%}
**Part 2**

```{r}
#| echo: true
#| eval: false

## Open GBIF occurrences ----

gbif_occ <- read.csv("gbif_occurrences.csv")


## Extract & clean species names ----

gbif_species <- gbif_occ |> 
  
  select(scientificName) |> 
  
  distinct() |> 
  
  mutate(scientificName = stringr::str_extract(scientificName, 
                                               "^[A-z]+\\s[a-z]+")) |> 
  
  rename(binomial_name = scientificName)

```

<br/>

**Part 3**

```{r}
#| echo: true
#| eval: false

## Merge tables ----

final_data <- left_join(gbif_species, pantheria, by = "binomial_name")
```
:::

::::
